---
title: "Support Vector Machines"
author: "Dean Kilfoyle"
date: "6 June 2016"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Support Vector Machines

```{r}
## Step 2: Exploring and preparing the data ----
# read in data and examine structure
letters <- read.csv("chapter07/letterdata.csv")

# divide into training and test data
letters_train <- letters[1:16000, ]
letters_test  <- letters[16001:20000, ]
```


## Step 3: Training a model on the data ----
# begin by training a simple linear SVM
library(kernlab)
letter_classifier <- ksvm(letter ~ ., data = letters_train,
                          kernel = "vanilladot")

# look at basic information about the model
letter_classifier

## Step 4: Evaluating model performance ----
# predictions on testing dataset
letter_predictions <- predict(letter_classifier, letters_test)

head(letter_predictions)

table(letter_predictions, letters_test$letter)

# look only at agreement vs. non-agreement
# construct a vector of TRUE/FALSE indicating correct/incorrect predictions
agreement <- letter_predictions == letters_test$letter
table(agreement)
prop.table(table(agreement))

## Step 5: Improving model performance ----
set.seed(12345)
letter_classifier_rbf <- ksvm(letter ~ ., data = letters_train, kernel = "rbfdot")
letter_predictions_rbf <- predict(letter_classifier_rbf, letters_test)

agreement_rbf <- letter_predictions_rbf == letters_test$letter
table(agreement_rbf)
prop.table(table(agreement_rbf))
