---
output: html_fragment
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
library(ggplot2)
library(magrittr)
# render this file with knit("neuralnet/math.Rmd")
```

### Procedure

1. Initialize

  1. Grid to width * height * dimensions of input
  2. Starting neighbor width to max(gridWidth,gridHeight) / 2
  3. Starting learning rate to 0.05
  4. Weights random -1..1
  
2. Prepare training data

  1. Scale(data) to mean centred on 0 and ideally range -1 to 1

3. Loop per iteration

  1. Using inputs find the Best Matching Unit based on euclidean distance

$$ \sqrt{\sum_{i=1}^n(p_i-q_i)^2} $$

  2. Update weights in the region of the BMU (v) using scaling functions theta and alpha where theta is the nighborhood function and alpha is the per iteration learning rate.

$$ W_v(t+1) = W_v(t) + \theta(v,t) \times \alpha(t) \times (X(t) - W_v(t)) $$

  3. Use the Gaussian function as a radial basis function for neighbour weighting (x1,x2 are the row and column disctnaces from the BMU)

$$ \theta(x_1,x_2) = e^{-v} $$

$$ v = \sum_{i=1}^n\frac{(x_i-c)^2}{2w^2} $$

```{r}
dist = 1^2:20^2 + 1^2:20^2
theta_d = exp(-dist/(2*10^2))
plot(theta_d)
```

  4. Ideally the width (w) and learning rate alpha should decay exponentially
  
  $$ \sigma(t) = \sigma_0 e^{-\frac{t}{\lambda}} $$
  
  
  where $\lambda = \frac{t}{maxt}$
  


```{r}
sigma_0 = 40/2
lambda = 100 / log(sigma_0)
sigma_t = sigma_0 * exp(-(1:100)/lambda)
plot(sigma_t)
```


